---
title: LearnSeurat_CITE_seq
author: Landau1994
avatar: /images/faceicon.png
authorLink: 'https://github.com/Landau1994'
authorAbout: 'https://github.com/Landau1994'
authorDesc: A PhD student in bioinformatics
mathjax: true
categories:
  - implementation
tags:
  - R
date: 2020-04-20 
keywords:
  description:
photos:
output:
  md_document:
    variant: gfm
always_allow_html: true
---

### 前言 

CITE-seq是Rahul Satija和Peter Smibert两个组合作开发的在单细胞精度，同时测量细胞表面蛋白表达和转录组的[技术](https://www.nature.com/articles/nmeth.4380)。该技术原理如下：

[CITE-seq 原理图，来自原始文献]()

该项技术可以用于免疫相关的单细胞测序研究中。例如, 有[研究表明](https://www.nature.com/articles/s41586-020-2134-y)称：

> 他们在人和小鼠非小细胞肺癌中进行单细胞RNA测序，鉴定了一群DC，并将其命名为“富含免疫调节分子的成熟DC”（mregDC），这是由于它们共表达了免疫调节基因（Cd274，Pdcd1lg2和Cd200）和成熟基因（Cd40，Ccr7和Il12b）。

这段中文报道来自[小柯机器人](http://news.sciencenet.cn/htmlpaper/2020/3/20203301537131655622.shtm)

Rahul Satija组开发的软件Seurat有一个[教程](https://satijalab.org/seurat/v3.1/multimodal_vignette.html)，可以分析CITE-seq数据。本文基于该教程对该类型数据的分析进行说明。

### 数据载入

首先我们需要获取数据，该数据集取样为8617个脐带血单核细胞，包含了表达谱数据和11个抗体来源标签数据（antibody-derived tags ,ADT)。

```{r,message=FALSE,warning=FALSE}
library(Seurat)
library(SeuratData)
library(ggplot2)
library(patchwork)
### AvailableData() check avaliable data: we choose cbmc
### InstallData('cbmc')
library(cbmc.SeuratData)
data("cbmc")
### expression matrix
cbmc[["RNA"]]@counts[1:10,1:10]
### ADT count matrix
### Actually there are just 10 surface protein
cbmc[["ADT"]]@counts[1:10,1:10] 
### show default assay
DefaultAssay(cbmc)
```

### 根据基因表达进行聚类

注意在默认参数的情况下，下述操作时对`Default Assay`进行的

```{r,fig.align="center",fig.height=6,message=FALSE,warning=FALSE}
# standard log-normalization
cbmc <- NormalizeData(cbmc)
# choose ~1k variable features
cbmc <- FindVariableFeatures(cbmc)
# standard scaling (no regression)
cbmc <- ScaleData(cbmc)
# Run PCA, select 13 PCs for tSNE visualization and graph-based clustering
cbmc <- RunPCA(cbmc, verbose = FALSE)
```

下面的图是根据标准差来选择PCs

```{r,fig.align="center",fig.height=6,message=FALSE,warning=FALSE}
ElbowPlot(cbmc, ndims = 50)
```

聚类和t-SNE降维

```{r,fig.align='center',fig.height=6,message=FALSE,warning=FALSE}
cbmc <- FindNeighbors(cbmc, dims = 1:25)
cbmc <- FindClusters(cbmc, resolution = 0.8)
cbmc <- RunTSNE(cbmc, dims = 1:25, method = "FIt-SNE")

# Find the markers that define each cluster, and use these to annotate the clusters, we use
# max.cells.per.ident to speed up the process
cbmc.rna.markers <- FindAllMarkers(cbmc, max.cells.per.ident = 100, min.diff.pct = 0.3, only.pos = TRUE)

# Note, for simplicity we are merging two CD14+ Monocyte clusters (that differ in expression of
# HLA-DR genes) and NK clusters (that differ in cell cycle stage)
new.cluster.ids <- c("Memory CD4 T", "CD14+ Mono", "Naive CD4 T", "NK", "CD14+ Mono", "Mouse", "B", 
    "CD8 T", "CD16+ Mono", "T/Mono doublets", "NK", "CD34+", "Multiplets", "Mouse", "Eryth", "Mk", 
    "Mouse", "DC", "pDCs")
names(new.cluster.ids) <- levels(cbmc)
cbmc <- RenameIdents(cbmc, new.cluster.ids)
```

我们看看聚类结果：

```{r,fig.align='center',fig.height=6,fig.width=6,message=FALSE}
DimPlot(cbmc, label = TRUE) + NoLegend()
```



### 蛋白表达数据处理

Seurat3的assay实现多个组学或者模态的数据的存储和获取。
代码里的注释来自Seurat官网。

```{r,message=FALSE,warning=FALSE}
# Now we can repeat the preprocessing (normalization and scaling) steps that we typically run
# with RNA, but modifying the 'assay' argument.  For CITE-seq data, we do not recommend typical
# LogNormalization. Instead, we use a centered log-ratio (CLR) normalization, computed
# independently for each feature.  This is a slightly improved procedure from the original
# publication, and we will release more advanced versions of CITE-seq normalizations soon.
cbmc <- NormalizeData(cbmc, assay = "ADT", normalization.method = "CLR")
cbmc <- ScaleData(cbmc, assay = "ADT")
```

在RNA表达谱的降维Embedding中同时展示展示蛋白表达水平和基因表达水平：

散点图：横纵轴为降维的坐标：

```{r,fig.align='center',fig.width=6,fig.height=8,message=FALSE}
# in this plot, protein (ADT) levels are on top, and RNA levels are on the bottom
FeaturePlot(cbmc, 
            features = c("adt_CD3", "adt_CD11c", 
                         "adt_CD8", "adt_CD16", 
                         "CD3E", "ITGAX", "CD8A", "FCGR3A"),
            min.cutoff = "q05", 
            max.cutoff = "q95", 
            ncol = 2)
```

Ridge Plot:

```{r,fig.align='center',fig.width=8,fig.height=8,message=FALSE}
RidgePlot(cbmc, features = c("adt_CD3", "adt_CD8", "CD3E","CD8A"),ncol = 2)
```

散点图：横纵轴为表达量；这个类似于FACS

```{r,fig.align='center',fig.width=8,fig.height=8,message=FALSE}
# Draw ADT scatter plots (like biaxial plots for FACS). Note that you can even 'gate' cells if
# desired by using HoverLocator and FeatureLocator
FeatureScatter(cbmc, feature1 = "adt_CD19", feature2 = "adt_CD3")
```

我们也可以看看蛋白表达和基因表达的关系：

```{r,fig.align='center',fig.width=8,fig.height=8,message=FALSE}
# view relationship between protein and RNA
FeatureScatter(cbmc, feature1 = "adt_CD3", feature2 = "CD3E")
```

我们可以看看T细胞：

```{r,fig.align='center',fig.width=8,fig.height=8,message=FALSE}
# Let's plot CD4 vs CD8 levels in T cells
tcells <- subset(cbmc, idents = c("Naive CD4 T", "Memory CD4 T", "CD8 T"))
FeatureScatter(tcells, feature1 = "adt_CD4", feature2 = "adt_CD8")
```

选没有标准化的原始数据我们看看，坐标轴的间距太大，会有misleading

```{r,fig.align='center',fig.width=8,fig.height=8,message=FALSE}
# # Let's look at the raw (non-normalized) ADT counts. You can see the values are quite high,
# particularly in comparison to RNA values. This is due to the significantly higher protein copy
# number in cells, which significantly reduces 'drop-out' in ADT data
FeatureScatter(tcells, feature1 = "adt_CD4", feature2 = "adt_CD8", slot = "counts")
```

这里还是可以观察到dropouts现象的，据原作者说：
> If you look a bit more closely, you’ll see that our CD8 T cell cluster is enriched for CD8 T cells, but still contains many CD4+ CD8- T cells. This is because Naive CD4 and CD8 T cells are quite similar transcriptomically, and the RNA dropout levels for CD4 and CD8 are quite high. This demonstrates the challenge of defining subtle immune cell differences from scRNA-seq data alone.

画热图，Seurat3 加了 downsample的功能。

```{r,message=FALSE,warning=FALSE}
# Downsample the clusters to a maximum of 300 cells each (makes the heatmap easier to see for small clusters)
cbmc.small <- subset(cbmc, downsample = 300)
# Find protein markers for all clusters, and draw a heatmap
adt.markers <- rownames(cbmc.small[["ADT"]]@counts)

```

我们可以看看Seurat热图的默认配色（三个冒号可以看更为底层的函数）, 个人觉得并不好看。

```{r,fig.align='center',fig.height=6}
# using code from RColorBrewer to demo the palette
n = 200
par(mfrow=c(3,1))
image(
  1:n, 1, as.matrix(1:n),
  col = Seurat:::PurpleAndYellow(k=n),
  xlab = "PurpleAndYellow n", ylab = "", xaxt = "n", yaxt = "n", bty = "n"
)
image(
  1:n, 1, as.matrix(1:n),
  col = colorRampPalette(c("navy", "white", "firebrick3"))(n),
  xlab = "NavyWhite3Firebrick3 n", ylab = "", xaxt = "n", yaxt = "n", bty = "n"
)
image(
  1:n, 1, as.matrix(1:n),
  col = colorRampPalette(RColorBrewer::brewer.pal(11,"RdBu"))(n),
  xlab = "RdBu n", ylab = "", xaxt = "n", yaxt = "n", bty = "n"
)
```



把默认配色换掉,见
```{r,fig.align='center',fig.height=6}
mypal <- rev(colorRampPalette(RColorBrewer::brewer.pal(11,"RdBu"))(256))
#mypal2 <- colorRampPalette(c("navy", "white", "firebrick3"))(256)
DoHeatmap(cbmc.small, 
          features = unique(adt.markers), 
          assay = "ADT", 
          angle = 90,size = 3)+
  scale_fill_gradientn(colors  = mypal)
```

去除细胞杂质，
```{r,message=FALSE,warning=FALSE}
# You can see that our unknown cells co-express both myeloid and lymphoid markers (true at the
# RNA level as well). They are likely cell clumps (multiplets) that should be discarded. We'll
# remove the mouse cells now as well
cbmc <- subset(cbmc, idents = c("Multiplets", "Mouse"), invert = TRUE)
```

### 直接根据蛋白质表达水平进行聚类

```{r,message=FALSE,warning=FALSE}
# Because we're going to be working with the ADT data extensively, we're going to switch the
# default assay to the 'CITE' assay.  This will cause all functions to use ADT data by default,
# rather than requiring us to specify it each time
DefaultAssay(cbmc) <- "ADT"
cbmc <- RunPCA(cbmc, features = rownames(cbmc), reduction.name = "pca_adt", reduction.key = "pca_adt_", 
    verbose = FALSE)
```

再来看PCA(其实这里算是degenrate到线性组合了)

```{r,fig.align='center',fig.height=6}
DimPlot(cbmc, reduction = "pca_adt")
```

```{r,message=FALSE,warning=FALSE}
# Since we only have 10 markers, instead of doing PCA, we'll just use a standard euclidean
# distance matrix here.  Also, this provides a good opportunity to demonstrate how to do
# visualization and clustering using a custom distance matrix in Seurat
adt.data <- GetAssayData(cbmc, slot = "data")
adt.dist <- dist(t(adt.data))

# Before we recluster the data on ADT levels, we'll stash the RNA cluster IDs for later
cbmc[["rnaClusterID"]] <- Idents(cbmc)

# Now, we rerun tSNE using our distance matrix defined only on ADT (protein) levels.
cbmc[["tsne_adt"]] <- RunTSNE(adt.dist, assay = "ADT", reduction.key = "adtTSNE_")
cbmc[["adt_snn"]] <- FindNeighbors(adt.dist)$snn
cbmc <- FindClusters(cbmc, resolution = 0.2, graph.name = "adt_snn")

# We can compare the RNA and protein clustering, and use this to annotate the protein clustering
# (we could also of course use FindMarkers)
clustering.table <- table(Idents(cbmc), cbmc$rnaClusterID)
clustering.table
```

下面这个embeding 还是根据ADT来的（不过只要marker连续，只有10个也没有关系？）
```{r,fig.align='center',fig.height=6}
new.cluster.ids <- c("CD4 T", "CD14+ Mono", "NK", "B", "CD8 T", "NK", "CD34+", "T/Mono doublets", 
    "CD16+ Mono", "pDCs", "B")
names(new.cluster.ids) <- levels(cbmc)
cbmc <- RenameIdents(cbmc, new.cluster.ids)

tsne_rnaClusters <- DimPlot(cbmc, reduction = "tsne_adt", group.by = "rnaClusterID") + NoLegend()
tsne_rnaClusters <- tsne_rnaClusters + ggtitle("Clustering based on scRNA-seq") + theme(plot.title = element_text(hjust = 0.5))
tsne_rnaClusters <- LabelClusters(plot = tsne_rnaClusters, id = "rnaClusterID", size = 4)

tsne_adtClusters <- DimPlot(cbmc, reduction = "tsne_adt", pt.size = 0.5) + NoLegend()
tsne_adtClusters <- tsne_adtClusters + ggtitle("Clustering based on ADT signal") + theme(plot.title = element_text(hjust = 0.5))
tsne_adtClusters <- LabelClusters(plot = tsne_adtClusters, id = "ident", size = 4)

# Note: for this comparison, both the RNA and protein clustering are visualized on a tSNE
# generated using the ADT distance matrix.
wrap_plots(list(tsne_rnaClusters, tsne_adtClusters), ncol = 2)
```
对于该结果，作者是这么解释的：

> The ADT-based clustering yields similar results, but with a few differences
+ Clustering is improved for CD4/CD8 T cell populations, based on the robust ADT data for + CD4, CD8, CD14, and CD45RA
+ However, some clusters for which the ADT data does not contain good distinguishing protein markers (i.e. Mk/Ery/DC) lose separation You can verify this using FindMarkers at the RNA level, as well

### 更多

pbmc 10k的细胞也提供了CITE-seq的多模态数据，具体细节，请看Seurat官方[教程](https://satijalab.org/seurat/v3.1/multimodal_vignette.html)。

